---
title: "Investigating the Importance of Early Objective Control in Professional League of Legends"
subtitle: "Studying the Impact of Neutral Objectives on Match Outcomes"
author: 
  - Hritik Shukla
thanks: "Code and data are available at: https://github.com/hritikshuklas/earlyobjctrl_prolol"
date: today
date-format: long
abstract: "This paper investigates the importance of early control over neutral objectives in professional League of Legends (LoL) esports. Leveraging data from Tier 1 leagues worldwide, we analyze the impact of securing neutral objectives such as the Rift Herald, Elemental Drakes, and Baron Nasher on a team's odds of winning. Through logistic regression modeling, our findings reveal that early control of neutral objectives significantly increases a team's odds of winning, with the magnitude of this effect varying across different objectives. We also explore the debate between prioritizing Rift Herald or Elemental Drakes, shedding light on strategic choices in high-level LoL gameplay."
format: pdf
output:
  pdf_document: default
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(rstanarm)
library(arrow)
library(ggplot2)
library(knitr)
library(modelsummary)
```

```{r}
#| include: false
#| warning: false
#| message: false

raw_data <- read_csv("../data/raw_data/2023_LoL_esports_match_data_from_OraclesElixir.csv")

analysis_data <- read_parquet("../data/analysis_data/cleaned_data.parquet")

raw_data <- raw_data |>
  select("league","teamname","result","firstherald",
         "firstdragon","firstbaron") 

analysis_data <- analysis_data |>
  mutate(
    result = factor(
      result,
      levels = c("Loss", "Win")
    ),
    firstherald = factor(
      firstherald,
      levels = c("No", "Yes")
    ),
    firstdragon = factor(
      firstdragon,
      levels = c("No", "Yes")
    ),
    firstbaron = factor(
      firstbaron,
      levels = c("No", "Yes")
    )
  )

model <- readRDS(file="../models/neutral_obj_to_result_model.rds")
```

# Introduction {#sec-intro}

League of Legends has had one of the most popular esports scene in the
history of gaming. The 2023 LoL World Championship held in South Korea
had the highest viewership of any esports event ever, with viewership
peaking at 6.4 million concurrent viewers [@worldsviewership]. However,
at the same time, many viewers have voiced their opinions about the
current state of the game. Even though neutral objectives such as the
Rift Herald, Elemental Drakes as well as the iconic Baron Nasher are
important to the game's structure and strategies implemented by teams
across the world, many viewers are claiming that the meta has become
stale and boring because of the objective focused gameplay many teams
have adapted, leading to slow and calculated gameplay instead of the
volatile and explosive gameplay people tune in for [@objfocusedboring].

This paper aims to find out if at the highest levels of play, focusing
on these objectives matter at all or are pro players fighting for
negligible advantages. In particular, the focus of this paper will be on
early control of these objectives. By analysing games played in Tier 1
League championships around the world, our aim is to see if a team's
control on the first spawn of these neutral objectives influences their
odds of winning the match. We want to determine how big, if any, do
these objectives influence a team's ability to win a game.

Another decision that players have to make in the early game is whether
they want to prioritize the Rift Herald over Elemental Drakes. Both
spawn around the same time on the map but on the opposite ends, leading
to teams often needing to give one up for the other. This has been a
popular subject of debate amongst the community for years
[@riftordragon]. By finding out how each of these objectives effect the
outcome of the game, we aim to see which objective teams should focus on
between the two.

The remainder of this paper is structured as follows:

-   @sec-data explores the dataset and the variables within it used for
    the study

-   @sec-model explains our models setup and our assumptions going into
    the study

-   @sec-results explores our findings in detail

-   @sec-discussion evaluates these findings in the context of our
    reality

I used R [@citeR], along with multiple packages to aid in the data
analysis and modelling. More particularly, the packages tidyverse
[@tidyverse], dplyr [@citeDplyr], tidyr [@citeTidyr] and arrow
[@citeArrow] were used for data acquisition, testing and cleaning. The
package rstanarm [@citeRstanarm] was used for modelling, and ggplot2
[@citeGgplot2], knitr [@citeKnitr] and modelsummary [@citeModelsummary]
were used for data visualization.

# Data {#sec-data}

The data used for this paper was acquired from Oracle's Elixir, founded
in 2015 by Tim Sevenhuysen, a veteran in the esports Industry. Data from
Oracle's Elixir is used by coaches, analysts and other professionals in
the industry. This dataset keeps track of most professional leagues for
League of Legends, major or minor, from all around the world and
contains in-depth player, team and champion analytics
[@oracleselixirabout]. This data is collected from various sources, such
as the official LoL esports website [@lolesportsweb], official websites
for individual leagues, fan-maintained wikis, Riot APIs, etc. We aim to
study the influence of neutral objectives on the result of a game in
Tier 1 leagues around the world. More specifically, we want to observe
if being the first one to capture and total number of captures of the
Rift Herald, Elemental Drakes, Baron Nashers, and Elder Drakes helps or
hinders the chances of a team winning.

## Data Cleaning {#sec-data-cleaning}

```{r}
#| label: tbl-raw-data-preview
#| tbl-cap: Raw Dataset Preview
#| echo: false

kable(head(raw_data, 10), booktabs=TRUE, format="latex")
```

The "league ", "split ", "role", "teamname", "side", "result",
"firstherald", "heralds", "firstbaron", "barons", "firstdragon",
"elementaldrakes", "elders" variables from CCES 2020 dataset (as seen in
@tbl-raw-data-preview) were selected initially.

The dataset contains match data for each individual player as well as
for the team as a whole. We are interested in the performance of the
team as a whole, so we filter out any individual player data using the
"role" attribute.

Next, we filter out any matches that didn't occur in a Tier 1 League,
since we aim to study the influence of neutral objectives at the highest
levels of play. The Tier 1 Leagues in League of Legends are:

-   LCS, League Championship Series (North America)

-   LEC, The League of Legends EMEA Championship (EMEA region)

-   LCK, League of Legends Champions Korea (South Korea)

-   LPL, League of Legends Pro League (China)

-   VCS, Vietnam Championship Series (Vietnam)

-   PCS, Pacific Championship Series (Taiwan)

-   CBLOL, Campeonato Brasileiro de League of Legends (Brazil)

-   LLA, Liga Latinoamérica (Latin America)

Any entries which were missing any of the these variables we needed were
removed during cleaning. This omission culled down our total match
entries from 125,904 to 4,260. However, we still have more than enough
data to derive meaningful results.

Finally, the values for the Boolean categorical variables -
"firstherald", "firstdragon", "firstbaron" - within the dataset were
cleaned to more human-readable values. In the raw dataset, 0 was used to
represent a non-positive outcome and 1 was used to represent a positive
outcome. For example, for the "result" variable, "0" represented loss
and "1" represented win; for the "firstherald" variable, "0" represented
false and "1" represented true. In the cleaned dataset, these numbers
were replaced by their intended values.

The end result of the cleaning process leaves the dataset shown in
@tbl-cleaned-data-preview.

```{r}
#| label: tbl-cleaned-data-preview
#| tbl-cap: Cleaned Dataset preview
#| echo: false

col_names <- c("League", "Result", "Took First Herald","Took First Dragon", "Took First Baron")
kable(head(analysis_data, 10), format="latex", booktabs=TRUE, col.names=col_names)
```

## Rift Herald {#sec-data-herald}

```{r}
#| label: fig-herald-histogram
#| fig-cap: "First Herald Captured"
#| echo: false
#| fig-subcap: ["First Herald Captured", "First Herald Capture Split by Match Result"] 
#| layout-ncol: 2

# Create a histogram
ggplot(analysis_data, aes(x = as_factor(firstherald))) +
  geom_bar() +
  labs(x = "", y = "Number of Teams") +
  theme_minimal()

ggplot(analysis_data, aes(x = as_factor(firstherald))) +
  geom_bar() +
  labs(x = "", y = "Number of Teams") +
  facet_wrap(~ result, scales="free") +
  theme_minimal()
```

The Rift Herald is an early game neutral objective which spawns, up to
two times, 8 minutes into the game and despawns 20 minutes into the
game. It can be captured twice within this time span. Rift Herald, when
used correctly, can help snow ball a team's leads or help them catch up
to the enemy, as it helps take down enemy towers and provides with gold
very early into the game, which pro players can take advantage of to
build even greater leads.

From our data, we can see that the first herald capture is split evenly,
which is as expected as we take the team data of both teams in a match,
and only one of them can have the first capture on the herald. It is
also a possibility that neither teams contest for this objective,
however this scenario seems to be rare, if not completely missing from
our data. So, we can safely say that Rift Herald is an important
objective for teams in the early game, and is always captured during its
availability period.

When we split our data by match results, a trend appears - winning teams
had a higher frequency of capturing the herald first, whereas losing
teams had a higher frequency of not capturing the herald first. This
difference is large enough to be of statistical significance, and
requires more research into it.

## Elemental Drakes

```{r}
#| label: fig-drake-histogram
#| fig-cap: Elemental Drake Capture Statistics
#| echo: false
#| fig-subcap: ["First Drake Captured", "First Drake Capture Split by Match Result"] 
#| layout-ncol: 2

# Create a histogram
ggplot(analysis_data, aes(x = as_factor(firstdragon))) +
  geom_bar() +
  labs(x = "", y = "Number of Teams") +
  theme_minimal()

ggplot(analysis_data, aes(x = as_factor(firstdragon))) +
  geom_bar() +
  labs(x = "", y = "Number of Teams") +
  facet_wrap(~ result, scales="free") +
  theme_minimal()

```

The Elemental Drakes are a recurring neutral objective in this game that
spawn multiple times per game. The first elemental drake spawns 5
minutes into the game. After being captured, it spawns again 5 minutes
later, up until either of the two teams have four total captures
combined. The drakes offer gold to the members of the team upon kill, as
well as different buffs which depend its base element (Infernal, Cloud,
Ocean, etc.). If a team is first to capture four of the drakes, the team
also gains a more powerful buff on top of the buffs already given,
giving them a great advantage to help them secure a win.

As expected again, we can see in @fig-drake-histogram that the first
dragon capture is split evenly, which is as expected again since the
drakes are a high priority objective and drake stacking is considered a
powerful strategy towards winning the game.

When we split our data by match results, we see a similar split as we
saw with the Rift Herald - winning teams had a higher frequency of
capturing the herald first, whereas losing teams had a higher frequency
of not capturing the herald first. This difference is slightly larger
than the one we saw in @fig-herald-histogram .

## Baron Nasher {#sec-data-baron}

```{r}
#| label: fig-baron-histogram
#| fig-cap: Baron Nasher Capture Statistics
#| echo: false
#| fig-subcap: ["First Baron Captured", "First Baron Capture Split by Match Result"] 
#| layout-ncol: 2

# Create a histogram
ggplot(analysis_data, aes(x = as_factor(firstbaron))) +
  geom_bar() +
  labs(x = "", y = "Number of Teams") +
  theme_minimal()

ggplot(analysis_data, aes(x = as_factor(firstbaron))) +
  geom_bar() +
  labs(x = "", y = "Number of Teams") +
  facet_wrap(~ result, scales="free") +
  theme_minimal()

```

Baron Nasher is a late-game neutral objective which replaces the Rift
Herald. It is a powerful monster with a massive health pool which
usually requires the whole team to be present to be conquered, and is
often highly contested due to its bounty. The team who kills baron
successfully is rewarded with a ton of gold as well as a temporary buff
which helps them take down the enemy base easier, increasing the odds in
their favor heavily. Due to the nature of its buff, its generally
considered to be a game ending decision and often converts a significant
lead into a game ending push, but sometimes, although very rare, a
clever steal can also result in the comeback for a team who has fallen
behind.

Unlike the Rift Herald and the drakes, the first Baron data in
@fig-baron-histogram isn't evenly split. Although the difference is
quite small, there is a higher frequency of teams who didn't get the
first baron. This can only be explained if neither teams ever slayed the
Baron, which does happen in games which snowball early really hard and
end really quick. These games are quite uncommon, which is why the
difference appears to be small.

Finally, we see the true power of the Baron buff as a game-ender when we
split our data by match results. We see that almost all teams who won
their games got the first baron buff of the game, whereas almost all
losing teams never got the first baron buff of the game. There are way
too many game states to analyze in depth of how Baron can influence the
odds of winning - an already snowballing team with a massive lead takes
control of the Baron easily and conquers it with no hindrance from the
enemy team; the fight for the buff ends a stalemate, etc. Another thing
to note is that from the split, we can see that it is not a
requirement - there are some people who won the game without ever
getting the first Baron, but these graphs don't tell the whole story. As
explained before, the game could have been a stomp and the winning team
never needed the Baron in the first place, or the team made a comeback
and lasted long enough to win the second baron fight. The game states in
League of Legends can get as varied and complicated as chess, and it is
out of the scope of this paper to delve into it. However, it is an
interesting case to study further.

# Model {#sec-model}

We will be using logistic regression to model our data, where our
outcome variable would be whether a team wins or loses their match.
Whether a team secures the first herald, the first elemental drake, or
the first baron will be used as predictors our outcome variable.

## Model set-up

Define:

-   $y_i$ is the result of the match - whether a team won or lost

-   $\mbox{firstherald}_i$ describes whether the team secured the first
    Rift Herald spawn

-   $\mbox{firstdragon}_i$ describes whether the team secured the first
    Elemental Drake spawn

-   $\mbox{firstbaron}_i$ describes whether the team secured the first
    Baron Nasher spawn

```{=tex}
\begin{align}  
  y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\ 
  \mbox{logit}(\pi_i) &= \beta_0 + \beta_1 \times \mbox{firstherald}_i + \beta_2 \times \mbox{firstdragon}_i + \beta_3 \times \mbox{firstbaron}_i \\
  \beta_0 &\sim \mbox{Normal}(0, 2.5) \\ 
  \beta_1 &\sim \mbox{Normal}(0, 2.5) \\ 
  \beta_2 &\sim \mbox{Normal}(0, 2.5) \\
  \beta_3 &\sim \mbox{Normal}(0, 2.5) 
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of
@rstanarm. We use the default priors from `rstanarm`. All of our
variables are Boolean categorical variables, and therefore have been
converted to factors so the model can treat them as such.

### Model justification

Logistic regression is employed for this model since our variable of
interest is a binary outcome variable - a win or a loss. Logistic
regression is well suited for situations where the outcome variable
represents two mutually exclusive categories and its probability is
based on a set of predictor variables (listed above).

Since both Rift Herald and drakes are available early on in the game, it
would be interesting to see if early priority for both of these
objectives would hold much influence.

The Rift Herald should have a positive effect on the result of the game,
however I expect it to be the smallest of the three. It gives a direct
reward which is good for the short moment after its utilized, but this
advantage doesn't scale well into the late game if the player is not
able to use it to force a greater advantage for themselves.

It is expected for the first drake to have a greater influence on the
game state than the Rift Herald, but less than the first Baron Nasher.
The elemental drake buffs only get more powerful as you stack more and
more on top of each other, and it gets really difficult for the opposing
team to match this strength later on in the game. Dragon buff stacking
is a strategy that takes time to take effect but is a guaranteed way to
build a sizable advantage against your opponents, and stacking the buffs
from the first drake spawn can allow teams to gain this advantage
earlier than they would if they waited, and put pressure on the enemy
team to find ways from the drake stacking to continue. It is important
to note that it is a common occurrence in pro play to trade the Rift
Herald for an early Drake, so it will also be interesting to see which
of these two objectives teams should focus on early in the game.

Finally, it is expected that our last neutral objective, the Baron
Nasher, will have the greatest effect on the result of the game, much
greater than the herald and the drake. This is because the Baron spawns
much later in the game and is often one of the last buffs a team with a
lead aims for before making their game ending push.

# Results {#sec-results}

```{r}
#| label: tbl-model-summary
#| tbl-cap: Model Summary
#| echo: false
#| warning: false

modelsummary(
  list(
    "Win" = model
    ), 
  statistic="mad")

```

Note that for our model, we set "Loss" as our reference level for
"result", and "No" as our reference level for "First Herald", "First
Dragon", and "First Baron". This means that the coefficients of first
objective captures are relative to the game state where a team isn't the
first to capture them, which we have seen from the histograms tends to
be more frequent when a team loses. R appropriately treats "Loss" as our
failure case and "Win" as our success case, which directly corresponds
to our coefficients' signs. Negative coefficients imply a decrease in
odds for a win. Conversely, a positive coefficient imply an increase in
odds for a win. Keeping these key pieces of information in mind, we
proceed to our model's results.

From @tbl-model-summary, we can see that our model arrived at the
intercept of -2.168, with a standard error of 0.090. This means that we
can say with fairly high accuracy, that when all predictors are set to
0, i.e, when a team is not the first one to capture any of the neutral
objectives, a team's log odds of winning are -2.168, which translates to
roughly 10.27% probability of winning, using the formulas:

$$
odds = e^{log\_odds}   
$$

$$
p = \frac{odds}{1 + odds}
$$

As predicted, the coefficients are positive for each of these buffs.
However, what's interesting is the scale of these individual
coefficients. While they agree with our hypothesis, the difference in
their magnitudes is exceeds our expectations.

First Herald, as predicted, is the least influential neutral objective,
with the lowest coefficient of the three, coming at 0.485 with a
standard error of 0.086. This falls in line with our hypothesis as we
expected it to be the least influential objective. The first Drake is
much more influential in deciding the fate of a game than the Rift
Herald, with an increase of 0.903 in log odds when it is captured first,
and a standard error of 0.084. Finally, we see that capturing the first
Baron has the most substantial effect on the chances of a team winning,
with 3.097 increase in log-odds of winning, with a standard error of
0.083.

# Discussion {#sec-discussion}

## What was done {#sec-first-point}

Objectives are an important part of League of Legends, and are important
part of strategies adopted by teams in the professional scene of the
game. The objective of this paper was to see how much early control of
these objectives influenced a team's odds of winning in the Tier 1
professional leagues around the world. This was done by obtaining match
data for all Tier 1 professional matches held in 2023 from Oracle's
Elixir, a popular website among esport professionals, analysts and
journals. Initial exploratory data analysis suggested that teams who won
more often than not had early control of these objectives. This
difference was much greater when it came to Baron Nasher when compared
to both the Rift Herald and Drakes. Logistic regression modelling was
then used to formalize this discovery, revealing a clear trend that
aligned with our initial hypothesis: early control of Rift Herald,
Elemental Drakes, and Baron Nasher progressively enhances a team's odds
of winning, with the magnitude of this increase growing in the order
they are listed.

## Early Rift Herald Control Is Effective, but Not the Best Strategy

Early control of the Rift Herald is overall beneficial for teams, and
increases a team's odds of winning. It is however, not as great as one
might think. An increase of 0.485 log odds from the intercept value of
-2.168 increases a team's probability from 10.27% to 15.67%. This can be
explained by how Rift Herald works. It helps teams break down towers
faster in the early game and accelerate one of the players into a
greater advantage. It is up to this player then to use this edge to
propel other players on their team forward to make a bigger advantage
for the team as a whole. Failure to capitalize on this lead can result
in this lead diminishing over time - even if the player maintains this
lead for themselves, its value diminishes over time as the game goes on.
Worst case scenario, the enemy makes up for their deficit by just
playing well and any advantage gained from the Rift Herald is lost.
Therefore, while it does provide an advantage, it is not significant
enough compared to others.

## Early Drake Control Appears to be Better Than Rift Herald

Early Drake control increases a team's log odds of winning by 0.903. It
is not as high as one might expect, due to how Drakes work in League of
Legends. A team must keep stacking drakes to utilize their full
potential. Early Drake control can help a team get a head start towards
stacking or delay the opposition's strategy of Drake stacking. Early
Drake control can therefore hurt an opposition by delaying their plans
or by fast-tracking your own progress.

Early Drake control does prove to be more advantageous than the Rift
Herald, by almost double the log odds. Compared to the Rift Herald,
which only bumps up a team's chances of winning to 15.67%, getting early
drake control increases these odds to 22.01% - much more significant
than the Rift Herald. There are too many factors to account for to say
definitively that early drake control is objectively better than the
Rift Herald, however, it does seem so that for Season 13 of Tier 1
League of Legends, Early Drake control provided a greater advantage than
Early Herald control.

## Securing First Baron Nasher Often Leads to Victory

In @fig-baron-histogram, we saw how huge the discrepancy was in teams
who won who also secured the first Baron Nasher of the game. This falls
in line with what our model depicts. Securing the first Baron Nasher
overwhelmingly increases a team's odds for winning, when compared to
Rift Herald and Elemental Drakes, with a coefficient of 3.097. This
falls in line with our expectation that Baron Nasher has the most
significant impact on a team's chances of winning, however, the
difference magnitude of the coefficient compared to the two other
objectives is what's astonishing.

It is often the case that teams usually play around the Baron buff as a
means to close out a game, as it is great for sieging the enemy base.
However, I believe that this number might be skewed by other hidden
factors that haven't been accounted for in the model. For example, if a
team already has a sizable advantage before Baron even spawns, then yes,
getting this buff with push that advantage further, but it might not be
as big of a factor in their success as this coefficient makes it out to
be. More in-depth research is required to say for certain how much
exactly an early Baron capture influences the game, but this model gives
us an approximation of how the high odds of winning are for teams who
get to be the first ones to capture the Baron.

## Weaknesses and Next Steps

While we can make rough estimates as to how control of these neutral
objectives helps a team's odds of winning, it does not show the whole
picture. There are simply too many factors to account for to say for
certain if a strategy is good. Moreover, these strategies won't last
forever; League is constantly evolving, and metas might look very
different even in the same season of pro play.

League is a very complex game which evolves constantly, from new patches
that release every two weeks, changes to the map and structure of the
arena that players compete in, new items being introduced or reworked,
to new champions being added to the game as well as champion reworks.
Riot Games make balance changes to the game every two weeks, which
influences the overall meta of the game as the season progresses through
the year. Strategies that work today may not work tomorrow. Games played
even a week apart may not be playing under the same meta anymore - there
simply too many changes which forces teams to adapt their play style or
completely change them.

Other than the game itself, the skill of each individual player and how
the players work together as a team also changes how a team approaches
the game. A team might exploit a single person on the opposing team to
gain an advantage, or play a strategy that counters their opponents
strategy. Early game focused teams might ignore the drakes completely
and focus completely on the Rift Herald if it aligns with their play
style more - the game might end before the Baron even spawns. Teams who
play for the late game might emphasize drake stacking more than others
and might ignore the first Baron spawn as their team hasn't finished
powering up yet. Moreover, these strategies might change for the same
team from game to game.

Moving forward, it is important to recognize that League of Legends is
always changing, and there are many factors that affect how teams play
and win. We'll need to keep studying and learning to understand it
better.

\newpage

\appendix

# Appendix {.unnumbered}

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check. In @fig-ppcheckandposteriorvsprior-2 we compare the posterior
with the prior.

```{r}
#| eval: true 
#| echo: false 
#| message: false 
#| warning: false 
#| label: fig-ppcheckandposteriorvsprior 
#| layout-ncol: 2 
#| fig-cap: Examining how the model fits, and is affected by, the data 
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(model) + 
  theme_classic() + 
  theme(legend.position ="bottom")

posterior_vs_prior(model) + 
  theme_minimal() + 
  scale_color_brewer(palette = "Set1") + 
  theme(legend.position ="bottom") + 
  coord_flip()
```

## Diagnostics

```{r}
#| label: fig-model-cred-interval 
#| fig-cap: Credible Intervals
#| echo: false 
#| warning: false

#modelplot(model, conf_level = 0.9) + labs(x = "90 per cent credibility interval")
```

```{r}
#| echo: false 
#| eval: true 
#| message: false 
#| warning: false
#| label: fig-model-trace-rhat 
#| fig-subcap: ["Trace plot", "Rhat plot"] 
#| layout-ncol: 2

plot(model, "trace")

plot(model, "rhat")

```

\newpage

# References
